Predicting computational reproducibility of scientific pipelines using collaborative filtering

### 2018-07-12 paper.tex
Wrote II.A

% packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsfonts} % for R symbol (the set of real numbers)
\usepackage{color}
\usepackage[pdftex]{graphicx}
\usepackage{graphicx}
\usepackage[bookmarks=false]{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=black,filecolor=black,urlcolor=blue}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{stmaryrd} % for llbracket and rrbracket
\usepackage{subcaption}
\usepackage{nicefrac}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

@@

We consider the problem of predicting the values of a variable 
controlled by a hidden, parametrized, time dependent, deterministic 
process over time. We assume that the variable values is observed only for 
a limited number of time points, and we aim at predicting the values in the remaining time points.

@@

\section{Method}
To evaluate the reproducibility of a pipeline, a retrieved matrix from 
the subjects and file-ids is built. Subjects and file-ids are equivalent 
to users and items, respectively, as they are known in Collaborative Filtering (CF) technique.
Filling in the missing entries of this matrix encompasses some issues 
which take some barriers for use of Collaborative filtering technique as it 
has being used in e-commerce applications.
For instance, files-ids (items) are produced in a specific order in the 
pipeline however items are sequentially independent in general use of CF in e-commerce.
This issue adds constraints on the selection of CF training set. Despite of random 
selection strategy which does not taking care of item order, we had to come up with those 
selection strategies that respect the pipeline fundamental fact about file order generation.  
The aforementioned matrix (utility matrix), con be fully completed by processing all subjects 
in pipeline however the whole processing can takes loads of time. This possibility of 
having a potential complete populated matrix means despite of potential sparse matrices 
in e-commerce case studies here the utility matrix is not sparse and 
we can decide precisely which samples we need to select as our training set selection. 
In this regard, to avoid the cold start issue we can take the first file-id of all 
subjects in order to get sure that for all subjects there is at least one 
selected element in training ratio and in the same way, all generated files 
from the process of a random subject which can guarantee that all files 
are presented in the selection set at least for once.

@@

\subsection{Collaborative filtering using ALS}

We consider the collaborative filtering method as implemented in Spark's MLlb.
% Summarize Koren, Bell and Volinsky: https://dl.acm.org/citation.cfm?id=1608614

>>>>

% packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsfonts} % for R symbol (the set of real numbers)
\usepackage{color}
\usepackage[pdftex]{graphicx}
\usepackage{graphicx}
\usepackage[bookmarks=false]{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=black,filecolor=black,urlcolor=blue}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{stmaryrd} % for llbracket and rrbracket
\usepackage{subcaption}
\usepackage{nicefrac}
\usepackage{amsmath}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

@@

We consider the problem of predicting the values of a variable 
controlled by a hidden, parameterized, time dependent, deterministic 
process. We assume that the variable values are observed only for 
a limited number of time points, and we aim at predicting the values in the remaining time points.

@@

\section{Method}

To evaluate the reproducibility of a pipeline, a retrieved matrix from 
the subjects and file-ids is built. Subjects and file-ids are equivalent 
to users and items, respectively, as they are known in Collaborative Filtering (CF) technique.
Filling in the missing entries of this matrix encompasses some issues 
which take some barriers for use of Collaborative filtering technique as it 
has being used in e-commerce applications.
For instance, files-ids (items) are produced in a specific order in the 
pipeline however items are sequentially independent in general use of CF in e-commerce.
This issue adds constraints on the selection of CF training set. Despite of random 
selection strategy which does not taking care of item order, we had to come up with those 
selection strategies that respect the pipeline fundamental fact about file order generation.  
The aforementioned matrix (utility matrix), con be fully completed by processing all subjects 
in pipeline however the whole processing can takes loads of time. This possibility of 
having a potential complete populated matrix means despite of potential sparse matrices 
in e-commerce case studies here the utility matrix is not sparse and 
we can decide precisely which samples we need to select as our training set selection. 
In this regard, to avoid the cold start issue we can take the first file-id of all 
subjects in order to get sure that for all subjects there is at least one 
selected element in training ratio and in the same way, all generated files 
from the process of a random subject which can guarantee that all files 
are presented in the selection set at least for once.

@@

\subsection{Collaborative filtering}

Collaborative filtering is a technique to predict unknown values of a 
matrix called ``utility matrix" from the known ones. Traditionally, the 
matrix represents the ratings of items, represented by columns, by 
users, represented by rows. Ratings might be explicit, when users input 
ratings through a dedicated system, or implicit, when users' behaviors 
are analyzed to estimate their preference toward specific items. A 
complete introduction to collaborative filtering is provided 
in~\cite{mmdsbook}.

Several methods have been proposed to implement collaborative 
filtering. Item-item collaborative filtering~\ref{item-item} predicts 
the rating of item $i$ by user $u$ from the ratings of items similar to 
$i$ by user $u$. Likewise, user-user collaborative 
filtering~\ref{user-user} predicts the rating of item $i$ by user $u$ 
from the ratings of item $i$ by users similar to $u$. Both methods
have been used extensively for e-commerce applications.

A third class of methods, which we will apply to our problem, is based 
on the factorization of the utility matrix to estimate latent factors 
with which users and items are represented. This method is described 
in~\cite{koren} and became famous as it allowed its authors to win the 
Netflix prize in 2009. To summarize, the method aims at finding $q_i$ 
and $p_u$ vectors of $\mathbb{R}^f$, where $f$ is the number of latent factors, such that:
\begin{equation*}
r_{ui} = q_i^Tp_u,
\end{equation*}
where $r_{ui}$ is the rating of item $i$ by user $u$. In practice, the optimization
involves a regularization term to avoid overfiting particular users or items. The method
finds $q_i$ and $p_u$ that minimize the following objective:
\begin{equation*}
\sum_{(i,u) \in \mathbb{T}}\left( r_{ui} - q_i^Tp_u\right)^2+\lambda \left( \|{q_i}\|^2 + \|{p_u}\|^2\right)
,
\end{equation*}
where $\lambda$ is the regularization parameter and $\mathbb{T}$ is the training set.

It is also common to include user biases $b_u$ and item biases $b_i$ in 
the optimization, defined as the average deviation of user $u$ and item 
$i$ to the global average $\mu$. The goal then becomes to find $q_i$ and $p_u$ that minimize the 
following objective:
\begin{equation*}
\sum_{(i,u) \in \mathbb{T}}\left( r_{ui} - \mu - b_u - b_i - q_i^Tp_u\right)^2+\lambda \left( \|{q_i}\|^2 + \|{p_u}\|^2 + b_u^2 + b_i^2\right)
\end{equation*}

Two good techniques to perform the optimization are stochastic gradient descent and alternating least squares (ALS). 
In the remainder of this paper, we use matrix factorization as implemented in 
Apache Spark's MLLib version 2.~\todo{xxx.}, which uses ALS.
